---
title: "Analysing averages of exponential distributions"
author: "FD"
date: "11 Mai 2016"
output: pdf_document
references:
- id:	wikipedia_exponentialdist
  title: Exponential distribution
  author:
    family: Wikipedia - Exponential distribution
  issued:
    year: 2016
  url : https://en.wikipedia.org/w/index.php?title=Exponential_distribution&oldid=716355350
  note : [Online; accessed 11-May-2016]
- id: wikipedia_clt
  author:
    family: Wikipedia - Central limit theorem
  title: Central limit theorem
  issued:
    year: 2016
  url: https://en.wikipedia.org/w/index.php?title=Central_limit_theorem&oldid=717005743
  note: [Online; accessed 11-May-2016]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This analysis is about averages of exponential distributions. The data is artificially generated. Each observation is the average over 40 random numbers drawn from an exponential distribution with $\lambda$ equal to 0.2. 1000 observations are used.

## Summary

It was found that the distribution of the averages of 40 random numbers drawn from an exponential distribution behaves (to a certain extend) like a normal distribution. This is, as expected, in accordance with the central limit theorem.

## Background

The  exponential distribution function is defined as: $\lambda e^{-\lambda x}$ and be 0 for $x<0$.

The expected value of the exponential distribution is $1/\lambda$ as well as the standard deviation.

The standard deviation of the mean of random variables is the standard deviation of the random variables normalized by the sample size: $\sigma(\bar{X})=\sigma/\sqrt{n}$. For the mean of 40 variables with a standard deviation of $1/\lambda$, $\lambda=0.2$ this yields:

```{r}
lambda<-0.2
sigma<-lambda
sigmaMuExp<-1/0.2/sqrt(40)
muExp<-1/lambda
sprintf("Expected mean: %f, expected standard deviation: %f", muExp, sigmaMuExp)
```

According to the central limit theorem, the mean of values drawn from a suitable distribution should be **approximately** normally distributed (you will see below, how good the approximation really is). Thus we would expect something normally distributed around $5$ with a standard deviation of approx. $0.8$.

## Simulations

The data were simulated by 1000 runs of randomly drawing 40 samples from the exponential distribution. The distribution of the observed means was calculated using the R built-in hist function.

```{r, results='hide'}
set.seed(42)
randomMeans=NULL
for (i in 1 : 1000) randomMeans = c(randomMeans, mean(rexp(n=40, rate = 0.2)))
h<-hist(randomMeans, plot = F)
df<-data.frame(x=h$mids, probability=h$density)
```

## Results 

### Statistics of observed means

The average and standard deviation of all the observations is:

```{r}
muObs=mean(randomMeans)
sigmaMuObs=sd(randomMeans)
sprintf("Observed mean: %f, observed standard deviation: %f", muObs, sigmaMuObs)
```

These values are quite close to the theoretical values of $5$ and $0.79$ respectively but do not match exactly. More observations (data not shown) improve the situation a little bit.

### Distribution of observed means

Based on the observed data, a gaussian model was fitted. The start point for the optimization was the mean according to $1/\lambda$, the expected standard deviation of the sample mean of approx. $0.79$ and an estimate for the value of the normal distribution around $0$.

```{r}
# See: http://stackoverflow.com/a/15883835
g = function(p){
  d = p[3]*dnorm(df$x,mean=p[1],sd=p[2])
  sum((d-df$probability)^2)
}
o<-optim(c(5,0.79,0.5),g)
```

The parameters obtained from the optimization are:

```{r}
sprintf(
  "Mean: %f, standard deviation: %f, scale: %f of the gaussian fit",
  o$par[1],
  o$par[2],
  o$par[3]
)
```

On the first glance, these values look quite good and are close to the expected values.

Finally, lets look at the graph:

```{r}
library(ggplot2)

est<-function(x) o$par[3]/o$par[2]/sqrt(2*pi)*exp(-1/2*(x-o$par[1])^2/o$par[2]^2)
df$est=est(df$x)

print(
  ggplot()+ geom_bar(data=df, aes(x=x, y=probability, colour="Data"), stat="identity")+
    geom_point(data=df, aes(x=x, y=est, colour="Gaussian fit"))+
    geom_point(data=df, aes(x=df$x, y=dnorm(df$x, mean = 5, sd=0.8),
                            colour="Normal distribution")) +
    theme(legend.position = "bottom")+
    scale_color_manual(values=c("#999999", "#E69F00", "#56B4E9"))+
    labs(title="Observed, fitted and normal distribution of means")
)
```

It looks quite normal, as the values of standard deviations and means and the central limit theorem suggests. Interestingly, a Kolmogorow-Smirnow-Test gives ambiguous results depending on the used sample (data not shown), indicating, that the distribution is not so normal like it looks like. A quantile plot shows the deviations especially at the tails (see appendix).

\newpage

## Apendix

Quantile plot of the mean distribution shows deviations especially at the tails.

```{r}
qqnorm(randomMeans, main="Quantile plot of means compared to normal distribution quantiles")
qqline(randomMeans)
```